---
title: "R Notebook"
output: html_notebook
---

# ECON 57: Introduction to Statistics and Probability for Economists with R Programming

## Lecture 1: Introduction to R Programming - Part I

### Goals
1. Provide working code for reference.
2. Illustrate concepts with economics-related examples.
3. Include exercises for in-class practice.

## Contents

### Installation and Setup
- Installing R, and RStudio
- Markdown, and R notebooks

## Basics of R Programming
### Variables and Data Types

#### Defining Variables

In R, you can assign values to variables using the assignment operator `<-`. Here are examples:

```{r}
# Note: do Alt+- to get <- 
x <- 5                    # integer
y <- 2.5                  # float
name <- "Economist"       # text
is_student <- TRUE        # Boolean 
```

#### Data Types

The type of data determines the types of operations we can apply. R supports various data types:

- **Numeric**: Includes both integers and floating-point numbers. Numeric data can be manipulated using arithmetic operators such as addition (+), subtraction (-), multiplication (*), division (/), and modulus (%), among others.

- **Character**: Represents strings of text. Strings can be combined using the concatenation operator (paste() or paste0()) and compared using relational operators like equality (==) or inequality (!=).

- **Logical**: Represents TRUE or FALSE values. Logical values can be combined using logical operators like AND (&), OR (|), and NOT (!), allowing you to create more complex conditions for decision-making and filtering.

```{r}
# Numeric variables
age <- 30
income <- 50000.75

class(age)
class(income)

# Numeric operations
age_next_year <- age + 1
double_income <- income * 2

# Character variable
country <- "Argentina"
class(country)

# Character operations
greeting <- paste("Hello, I am from", country)

# Logical variable
is_employed <- TRUE
is_equal <- country == "Argentina"
class(is_equal)

# Logical operations
eligible_for_retirement <- age >= 65
not_employed <- !is_employed
```

```{r}
# Exercise 1 (Numeric and Character Operations)

# Given variables
num_cars <- 15
price_per_car <- 25000
manufacturer <- "AutoCo"

# Task 1: Calculate the total cost of all the cars (quantity * unit cost) and store it in a variable called total_cost.

# Task 2: Create a message using the manufacturer variable to introduce the car company and the total_cost variable to mention the total cost of cars. Store this message in a variable named intro_message.

# Print the results
cat("Total cost of cars:", total_cost, "\n")
cat(intro_message, "\n")
```

```{r, echo=FALSE}
# Solution 1
# Given variables
num_cars <- 15
price_per_car <- 25000
manufacturer <- "AutoCo"

# Task 1: Calculate the total cost of all the cars (quantity * unit cost) and store it in a variable called total_cost.
total_cost <- num_cars * price_per_car

# Task 2: Create a message using the manufacturer variable to introduce the car company and the total_cost variable to mention the total cost of cars. Store this message in a variable named intro_message.
intro_message <- paste(manufacturer, "Cost report: \n", total_cost)


# Print the results
cat("Total cost of cars:", total_cost, "\n")
cat(intro_message, "\n")

```

#### Vector

Vector is the simplest data structure we can implement in R. A vector is a sequence of elements of the same data type. You create a vector by grouping its elements between parentheses preceded by the c() function that combines them.

```{r}
# Numeric vector
prices <- c(10.5, 20.3, 15.2, 12.7)
avg_price <- sum(prices)/length(prices)
print(avg_price)

# Character vector
fruits <- c("apple", "banana", "orange", "grape")

# Logical vector
bool_values <- c(TRUE, FALSE, TRUE, TRUE)

# Numeric vector operations
discount <- 0.15
discounted_prices <- prices * (1 - discount)

# Character vector operations
new_fruit <- "pear"
updated_fruits <- c(fruits, new_fruit)

# Logical vector operations
all_true <- all(bool_values)
any_false <- any(bool_values)
```

##### Indexing Vectors

You can access elements in a vector using indexing:

```{r}
# Accessing second element of prices vector
second_price <- prices[2]
print(second_price)

# Multiple indexing

## Accessing first three elements of fruits vector
first_three_fruits <- fruits[1:3]     # also called slicing
print(first_three_fruits)

## Accessing non-continuous indices with c()
odd_fruits <- fruits[c(2, 4)]
print(odd_fruits)

# Conditional indexing

## Accessing low prices
low_prices <- prices[prices < 20]
print(low_prices)

print(prices < 20)
```

##### Modifying Vectors

You can modify vector elements using indexing as well:

```{r}
# Modify the second element of prices vector
prices[2] <- 22.0
print(prices)

# modify the second and third elements
prices[2:3] <- c(44, "gdp")
print(prices)
```

```{r}
# Exercise 2: Vector Operations (Calculating Total Revenue and Adding New Elements)

# Given numeric vector
daily_sales <- c(120.5, 89.3, 150.2, 105.7, 135.8)

# Given character vector
products <- c("Widget A", "Widget B", "Widget C", "Widget D", "Widget F")

# Task 1: Calculate the total revenue generated from the daily sales and store it in a variable named total_revenue. Print the total revenue and the number of days


# Task 2: A new product "Widget E" has been introduced and 36.7 million in revenue was generate its first day. Add this new product to the products vector and store the result in a variable named updated_products. Then, do the same for daily sales revenue. Add the new product in the appropriate index so it is in alphabetical order.


# Task 3: Update total revenue and store in new variable


# Print the results


```

```{r, echo=FALSE}
# Solution 2

# Given numeric vector
daily_sales <- c(120.5, 89.3, 150.2, 105.7, 135.8)

# Given character vector
products <- c("Widget A", "Widget B", "Widget C", "Widget D", "Widget F")

# Task 1: Calculate the total revenue generated from the daily sales and store it in a variable named total_revenue. Print the total revenue and the number of days
total_revenue <- sum(daily_sales)
num_days <- length(daily_sales)
paste("The cumulative revenue over the last", num_days, " days is ", total_revenue)

# Task 2: A new product "Widget E" has been introduced and 36.7 million in revenue was generate its first day. Add this new product to the products vector and store the result in a variable named updated_products. Then, do the same for daily sales revenue. Add the new product in the appropriate index so it is in alphabetical order.
new_product <- "Widget E"
updated_products <- c(products, new_product)
updated_products[6] <- updated_products[5]
updated_products[5] <- new_product

updated_daily_sales <- c(daily_sales, 36.7)

# Task 3: Update total revenue and store in new variable
updated_total_revenue <- sum(updated_daily_sales)

# Print the results
cat("Total revenue:", updated_total_revenue, "\n")
cat("Updated products:", updated_products, "\n")

```

##### Vector Operations

R supports element-wise vector operations:

```{r}
# Adding two vectors element-wise
vector1 <- c(1, 2, 3)
vector2 <- c(4, 5, 6)

summing <- vector1 + vector2
print(summing)

substracting <- vector1 - vector2
print(substracting)

multiplication <- vector1 * vector2
print(multiplication)
```

#### Factors: Handling Categorical Data

In addition to numeric, character, and logical data types, R also provides a special data type called `factor` to handle categorical data more efficiently. Factors are used to represent categorical variables and their levels. We won't go into detail on categorical data in this course, but I figured to include this in case you would like to use it for final project.

```{r}
# Creating a factor (i.e., dummy variable)
gender <- factor(c("Male", "Female", "Male", "Female", "Male"))
print(gender)

# Getting the levels of a factor
levels(gender)
```

Factors are useful because they represent categories more intuitively and are often used in statistical analyses. R also assigns an integer value to each level behind the scenes, making computations faster and more memory-efficient.

```{r}
# Creating a factor with specified levels
education <- factor(c("High School", "College", "High School", "PhD"),
                    levels = c("High School", "College", "Master's", "PhD"))
print(education)

# Checking the levels of a factor
levels(education)

# Converting a factor to character
education_char <- as.character(education)
print(education_char)
```

```{r}
# Simulating a dataset of student majors
majors <- c("Economics", "Computer Science", "Biology", "Economics", "Mathematics")

# Creating a factor from the majors vector
major_factor <- factor(majors)

# Analyzing the distribution of majors
major_distribution <- table(major_factor) # the table() command returns a frequency table (more on this next week)
print(major_distribution)
```

### Conditionals

```{r}
# If statement
x <- 10; y <- 20
if (x > y) {
  print("x is greater than y")
} else {
  print("x is not greater than y")
}

# Economics example: Buying decision
price <- 15
budget <- 20

if (price <= budget) {
  print("Affordable")
} else {
  print("Too expensive")
}

```

```{r}
# If-else if-else ladder
grade <- 75
if (grade >= 90) {
  print("A")
} else if (grade >= 80) {
  print("B")
} else if (grade >= 70) {
  print("C")
} else {
  print("D or below")
}

# Economics example: Income Tax Bracket
income <- 60000

if (income <= 50000) {
  tax <- income * 0.10
} else if (income <= 100000) {
  tax <- 50000 * 0.10 + (income - 50000) * 0.20
} else {
  tax <- 50000 * 0.10 + 50000 * 0.20 + (income - 100000) * 0.30
}

print(paste("Income Tax:", tax))
```

#### Logical Operators

R provides logical operators to combine conditions:

- &: AND operator (both statements true)

- |: OR operator (one of the statements true)

- !: NOT operator (negation)

```{r}
# Logical operators
temperature <- 25
is_raining <- FALSE

if (temperature > 20 & !is_raining) {
  print("Great day for a picnic!")
} else {
  print("Maybe another day...")
}

```


### Exercises

1. Create a variable age and assign your age to it. Print a message saying "You're young!" if your age is less than 30, otherwise, print "You're wise!"

```{r, echo=FALSE}
# Exercise 1
age <- 25

if (age < 30) {
  print("You're young!")
} else {
  print("You're wise!")
}

```


2. Create a variable that holds a vector that contains 10 elements of various data types. Print the 4th element and its type. Then, print elements 3 and 6 and 7 using conditional indexing.

```{r, echo=FALSE}
# Exercise 2
mixed_vector <- c(20, "hello", TRUE, 3.14, FALSE, "world", 42, 1.618, "R", -10)

# Print 4th element and its type
print(mixed_vector[4])
print(class(mixed_vector[4]))

# Print elements 3, 6, and 7 using conditional indexing
print(mixed_vector[c(3, 6, 7)])

```


3. Create a numeric variable to hold some number. Write a conditional statement to check if a given number is positive, negative, or zero. Print an appropriate message.

```{r, echo=FALSE}
# Exercise 3
number <- -5

if (number > 0) {
  print("The number is positive.")
} else if (number < 0) {
  print("The number is negative.")
} else {
  print("The number is zero.")
}

```


4. Create a factor named temperature with levels "Cold", "Moderate", and "Hot". Convert the factor to a character vector.

```{r, echo=FALSE}
# Exercise 4
temperature <- factor(c("Cold", "Moderate", "Hot"))

# Convert factor to character vector
temperature_char <- as.character(temperature)
print(temperature_char)

```


5. Create a program to calculate the cost of a hotel stay based on the following rules:

- Base cost per night: $100

- If the stay is longer than 7 nights, a discount of 10% is applied.

- If the stay is longer than 14 nights, a discount of 15% is applied.

```{r, echo=FALSE}
# Exercise 5
base_cost_per_night <- 100
stay_duration <- 8

if (stay_duration > 14) {
  discount_percent <- 15
} else if (stay_duration > 7) {
  discount_percent <- 10
} else {
  discount_percent <- 0
}

total_cost <- base_cost_per_night * stay_duration * (1 - discount_percent / 100)
print(paste("Total cost of the hotel stay: $", total_cost))

```


----
