---
title: "ECON 57: Economic Statistics - R Notebook 7"
output: html_notebook
---

# Probability in R

## Introduction

In R, we can perform probability analyses in two ways: 1) Applying the mathematical formulas we have covered in class, or 2) coding simulations 


```{r}
# Applying the formula P(A) = # favorable outcomes / sample size
# Sample space
S <- c(1, 2, 3, 4, 5, 6)

# Event A: rolling an even number
A <- c(2, 4, 6)

# Event B: rolling a number greater than 4
B <- c(5, 6)

# Compute the probabilities of A and B (assuming equally likely outcomes)
P_A <- length(A) / length(S)
P_B <- length(B) / length(S)

P_A
P_B
```

```{r}
# simple simulations
# Set the seed for reproducibility. Try running this cell with different seeds, what happens?
set.seed(123)

# Simulate a coin flip
coin_flips <- sample(c("Heads", "Tails"), size = 1, replace = TRUE)
print(coin_flips)
```

You can also calculate the probability of getting heads with a large number of coin flips

```{r}
# simulate 10000 coin flips. Run moew simulations varying the sample size and the sampling method. What happens as the sample size gets bigger and bigger?
coin_flips <- sample(c("Heads", "Tails"), size = 10000, replace = TRUE)
print(coin_flips)

# Calculate the probability of getting a head
prob_heads <- sum(coin_flips == "Heads") / length(coin_flips)
print(prob_heads)
```

Here, we have simulated flipping a fair coin 10,000 times and calculated the probability of getting a "head".

## II. Kolmogorov Axioms

Let's further illustrate these concepts using random experiments in R.

1. **Nonnegativity**: The probability of an event is always greater than or equal to 0.

Since R normalizes the probabilities to sum to 1 in the `sample()` function, we see that the chances of landing any event is always nonnegative.

```{r}
# Set the seed for reproducibility
set.seed(148)

# A 6-face fair die
die <- c(1, 2, 3, 4, 5, 6)

# Roll the die 1000 times
rolls <- sample(die, size = 1000, replace = TRUE)

# Probabilities (normalized frequencies) of each outcome
probs <- table(rolls) / length(rolls)
print(probs >= 0)
```

2. **Unit Measure**: The probability of the sample space is always 1.

We can see that the sum of all the probabilities comes up to 1.

```{r}
sum(probs)
```

3. **Additivity**: For any two mutually exclusive events, the probability that at least one of the events will occur is the sum of their individual probabilities.

Let's calculate the probability of landing 1 or 2.

```{r}
# Probability of landing a 1 or 2
prob_1_or_2 <- sum(probs[c("1", "2")])
print(prob_1_or_2)
```

## III. Conditional Probability 

Conditional probability is the probability of an event given that another event has occurred. Let's simulate a case where we flip two coins in a row. We're interested in knowing the probability of the second coin being heads given that the first coin was heads.

```{r}
# Set the seed for reproducibility
set.seed(111)

# Flipping a coin twice
coin_flips <- sample(c("Heads", "Tails"), size = 10000, replace = TRUE, prob = c(0.5, 0.5))

# Create two vectors: one for the first flip, and the other for the second flip
first_flip <- coin_flips[seq(from=1,to=10000,by=2)] # seq() is a built-in function to generate a sequence in R
second_flip <- coin_flips[seq(2,10000,2)]

# Find the probability that the second coin is a head given that the first coin was a head
# Apply the formula we learnt P(H|H) = P(H and H)/P(H)
# We count all the ocurrences in both flips in which both outcomes came out Heads, sum them, and divide them by the probability of getting Heads in any given roll
prob_second_head_given_first_head <- sum(first_flip == "Heads" & second_flip == "Heads") / sum(first_flip == "Heads")

print(prob_second_head_given_first_head)
```

In this code, we calculate the probability of drawing an Ace from a deck of cards, given that the previous draw was also an Ace.

# Class Activities 

1. **Decipher the Dice:** You roll a die 100 times. Calculate the probabilities of rolling a 1, a 2, a 3, a 4, a 5 and a 6. Also calculate the cumulative probability of rolling a 1, 2, or 3.

```{r}
# Set seed for reproducibility
set.seed(42)

# Roll a die 100 times
rolls <- sample(die, size = 100, replace = TRUE)

# Probabilities (normalized frequencies) of each outcome
probs <- table(rolls) / length(rolls)

# Cumulative probability of 1, 2 or 3
cumulative_prob <- sum(probs[c("1", "2", "3")])

probs
cumulative_prob
```

2. **Ace out the Cards:** Simulate drawing a card from a deck 5000 times. Calculate the following conditional probabilities: 
   - The chance of drawing a king given that the card drawn was a face card (Ace, King, Queen, and Jack are considered face cards)
   - The chance of drawing a face card given that the card drawn was a king.
   
```{r}
# Set the seed for reproducibility
set.seed(789)

# The 52 cards in a deck
cards <- c(rep("King", 4), rep("Face card", 12), rep("Not face card", 36))

# Draw a card 5000 times
draws <- sample(cards, size = 5000, replace = TRUE)

# Calculate conditional probabilities
prob_king_given_face <- sum(draws[1:(length(draws) - 1)] == "Face card" & draws[2:length(draws)] == "King") 
                         / sum(draws[1:(length(draws) - 1)] == "Face card")

prob_face_given_king <- sum(draws[1:(length(draws) - 1)] == "King" & draws[2:length(draws)] == "Face card") 
                          / sum(draws[1:(length(draws) - 1)] == "King")

prob_king_given_face
prob_face_given_king
```

3. **Economic Scenario: Stock Returns**

Suppose we have two stocks, stock A and stock B. The returns of both stocks are normally distributed. The return of stock A follows a normal distribution with mean 0.05 (5%) and standard deviation 0.1 (10%), while the return of stock B follows a normal distribution with mean 0.06 (6%) and standard deviation 0.2 (20%).

We can generate random returns for these two stocks and estimate the probability of negative return for each stock.

```{r}
# Set the seed for reproducibility
set.seed(234)

# Generate 10000 random returns for each stock
returns_A <- rnorm(10000, mean = 0.05, sd = 0.1)
returns_B <- rnorm(10000, mean = 0.06, sd = 0.2)

# Calculate the probability of negative return for each stock
prob_negative_return_A <- sum(returns_A < 0) / length(returns_A)
prob_negative_return_B <- sum(returns_B < 0) / length(returns_B)

print(paste0("Probability of negative return for stock A: ", prob_negative_return_A))
print(paste0("Probability of negative return for stock B: ", prob_negative_return_B))
```